---
title: "Exploring data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r panelset, echo=FALSE}
xaringanExtra::use_panelset()
```



## Gapminder data

In this lesson we will work with some of the data from the [Gapminder project](https://www.gapminder.org/). This data are available via an R package, which you will need to install in R.  
First, install the `gapminder` package. Then load the `gapminder` package and the `tidyverse` package as we will need its functionality to work with data.

```{r}
library(gapminder)
library(tidyverse)
```

The gapminder package provides you with a data table having the same name as the package (unfortunately).  
Let's inspect the contents of the data using the glimpse function from tidyverse:

```{r}
glimpse(gapminder)
```

The output tells us that the Gapminder data contain 1704 rows and 6 columns.  
An alternative to `glimpse()` is the `str()` function, showing the structure of the data

```{r}
str(gapminder)
```

## Handling data

If you have not already done so, install the tidyverse package now. Then, load it via:

```{r}
library(tidyverse)
```

Look at the data:

```{r}
gapminder
```

Look at the head (= top 6 rows) or tail (= last 6 rows) of the data

```{r}
head(gapminder)
tail(gapminder)
```


Inspect the variable names, the number of columns, the length of the data (= ncol), the dimension, and the number of rows:

```{r}
names(gapminder)
ncol(gapminder)
length(gapminder)
dim(gapminder)
nrow(gapminder)
```


You can get a quick statistical description of each variable with

```{r}
summary(gapminder)
```


## Look at some variables individually

Let's start by looking at some of the year values, and then count how many times each year appears in the data using the `table()` function:

```{r}
head(gapminder$year)
summary(gapminder$year)
table(gapminder$year)
```

```{r}
class(gapminder$continent)
summary(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
```


## Plotting

```{r}
ggplot(data = gapminder, aes(x = continent)) +
  geom_bar()
```

